Computing likelihoods under the SMC for a general class of ARGs.

### 2024-03-14 paper.tex
First pass at intro paragraph

\section{Abstract}
There currently is no way to connect the output of the diverse 
range of ARG inference methods back 
to a model that would allow us to perform a probabilistic exploration of the 
ARG space around the point estimate, or likelihood-based inference.

\section{Introduction}
\begin{itemize}
    \item ARGs are cool: There are now lots of ways to infer ARGs. Large scale either using heuristics or approximations.
    \item Models are good/essential: CwR useful but terrible for inference -> SMC
    \item Problem: we can't use models on inferred ARGs (Relate, tsinfer, ARGneedle). These methods produce point estimates. What we can't do:
        \begin{itemize}
            \item quantify uncertainty
            \item explore ARG space systematically around the point estimates. We don't have a metric on what an improvement might be.
            \item compare models of demography
            \item can't evaluate goodness of fit of model estimates
        \end{itemize}
    \item what we do here
\end{itemize}

[inspiration] great point from Wong et al.:
Note that it is important to distinguish here between structures and models: 
whether an inference method is based on 
heuristics or a rigorous mathematical model is orthogonal to the level 
of detail provided in its estimate. One could heuristically estimate a 
fully precise ARG, or statistically sample a partial, approximate ARG under 
a model such as the coalescent.

@@

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal 
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
provided here can function as a building block of [...] 
$N_e$-changes as well as multiple populations and migration can be readily incorporated 
in the current algorithm.\\

>>>>

\begin{abstract}
There currently is no way to connect the output of the diverse
range of ARG inference methods back
to a model that would allow us to perform a probabilistic exploration of the
ARG space around the point estimate, or likelihood-based inference.
\end{abstract}

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024}.
Additionally, there is a rich suite of tools growing around
the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
adrion_community_2020,terasaki_geonomics_2021,
baumdicker_efficient_2021,
korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
wohns_unified_2022,mahmoudi_bayesian_2022,
rasmussen_espalier_2022,
zhang_biobank-scale_2023,zhan_towards_2023} and analysis
packages
% etc
\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
Infererence methods now tend to fall into two classes: those that
sample from a statistical
model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022}
or parsimony~\citep{ignatieva_kwarg_2021} and that return
a fully precise ARG estimate~\citep{wong_general_2023},
and those based on
heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
zhang_biobank-scale_2023} that return an more approximate structure.
Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive grown in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.

\begin{itemize}
    \item Models are good/essential: CwR useful but terrible for inference -> SMC
    \item Problem: we can't use models on inferred ARGs (Relate, tsinfer, ARGneedle). These methods produce point estimates. What we can't do:
        \begin{itemize}
            \item quantify uncertainty
            \item explore ARG space systematically around the point estimates. We don't have a metric on what an improvement might be.
            \item compare models of demography
            \item can't evaluate goodness of fit of model estimates
        \end{itemize}
    \item what we do here
\end{itemize}

@@

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal
trees \citep{hejase_summary_2020} [TO DO: add more refs]. The theory and associated algorithm
provided here can function as a building block of [...]
$N_e$-changes as well as multiple populations and migration can be readily incorporated
in the current algorithm.\\
